---
title: "EFA in lavaan"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
options(scipen=999)

```

# lavaan (LAatent VAriable ANalysis)

Full documentation available at 
(https://cran.r-project.org/web/packages/lavaan/lavaan.pdf)

## Some Useful lavaan Notations

The "=~" operator can be used to define latent variables. This is to define a reflexive factor. 

The "<~" operator can be used to define a formative factor.

The "~~" (‘double tilde’) operator specifies variances, residual variances, and covariances.

The "|" operator can be used to define the thresholds of categorical endogenous variables.

As in Mplus, you can have multiple variables on the right side of the operator. If you do, you will need to add "+" in between them (if you list them without the "+" as you would in Mplus you will get an error message).

## lavaan Defaults to Know About

As in Mplus, defaults vary depending on the model being estimated, but they are not necessarily the same as in Mplus. We'll go over these as needed, but all this information is listed on the lavaan official documentation.  Many are identical to Mplus's defaults, but it's worth checking them first.

- Default estimator it maximum likelihood;
- The factor loading of the first indicator of a latent variable is fixed to 1;
- Residual variances are freely estimated;
- All exogenous variables are allowed to covary.

## Fixing covariances in lavaan

Building on the above, you can specify an orthogonal (zero) covariance between two latent or observed variables:
f1 ~~ 0*f2 

## If you have categorical indicators

We'll talk about categorical indicators later in the course. Muthen & Muthen recommend weighted least squares (WLS) when you have many factors and not so many factor indicators. They recommend maximum likelihood (ML, MLR) when you have few factors and many factor indicators. Both MLR and WLS can deal with categorical and continuous outcomes. Today, we'll use continuous indicators only.

## Mplus resources for EFA
### Not just code, the stats too

http://www.statmodel.com/discussion/messages/8/8.html

# Load packages 

```{r}
library(lavaan) # for the loadings
library(psych) # for EFA
library(apaTables) # for APA tables
library(dplyr) # for data manipulation
library(magrittr) # for the pipe operator
library(ggplot2) # for plots
library(semPlot) # for path diagrams
library(apaTables) # for APA tables

#psych::alpha()
```

# Load file

```{r}

library(haven) # for reading SPSS files
  efa_data <- read_sav("Data/IH Validation 1-21.sav") # command + shift + c to comment out

library(readr) # for reading csv files
  efa_data <- read_csv("Data/IH Validation.csv")

  efa_data <- read.table("Data/IH Validation 1-21 .dat")

  dim(efa_data) # check the dimensions of the data
  head(efa_data) # check the first few rows of the data
  tail(efa_data) # check the last few rows of the data
  psych::describe(efa_data) # get descriptives
  View(efa_data) # view the data in RStudio
  
```

# corr.test

```{r}

  corr.test(efa_data) # coefficient and p-values
  pairs.panels(efa_data) # scatterplots and histograms at once
  lowerCor(efa_data[,c("V1", "V2", "V3")]) # lower correlation plot 
  
  efa_data %>% # Lets get a scatterplot 
    ggplot(aes(x=V2, y=V5)) + 
    geom_point() +
    geom_jitter() +
    geom_smooth() + 
    theme_classic() 
  
  apa.cor.table(efa_data, "output/corr.doc") # APA table of correlations
  #apa.reg.table() , apa.aov.table()

```    

# EFA using fa in psych package

```{r}

  nfactors(efa_data, rotate = "oblimin") # get the number of factors
  
  efa_model = fa(efa_data, nfactors = 3, rotate = "oblimin", fm = "minres")
  
  print.psych(efa_model, digits = 2, sort = TRUE, cut = .4)

  efa_model %>% names # get the names of the output
  efa_model$loadings # get the loadings
  efa_model$communality # get the communality
  efa_model$uniqueness # get the uniqueness
  efa_model$fit # get the fit

  scree(efa_data, 
      factors = TRUE
      ) # get scree plot

```

# when you have missing data

```{r}
  # needs to be a matrix for this one so we'll practice making a 
  # matrix in which we use FIML to account for missing data
  # Full Information Maximum Likelihood (FIML) is a technique for handling missing data. Instead of doing listwise or pairwise deletion, it uses all the available information in the data to estimate the correlation (or covariance) among your variables.


  efa_matrix <- corFiml(efa_data, covar = FALSE, show = FALSE)

  # show = F means that you will do FIML, show = true means only showing missingness patterns (also   useful) but not doing FIML. covar = FALSE means we're getting a correlation (and not a covariance)   matrix
  

  # You can specify with columns to use without creating a new dataframe (e.g.,   corFiml(psychTools::efa_data[1:3],show = FALSE)). 

  #you can use efa_matrix in the fa() function as well


```

# Specify the model

The text ("block1") is a random label - you could call this latent variable anything you wanted, such as ("efa1").

This model defines two latent variables and tells lavaan that all six indicators (V1-V6) are allowed to load on both latent variables.

```{r}

# these are exploratory blocks (you can give them any name), you'll need to specify them
  efa_model <- '
  efa("block1")*F1 =~ V1 + V2 + V3 + V4 + V5 + V6
  efa("block1")*F2 =~ V1 + V2 + V3 + V4 + V5 + V6
  '
```

```{r}

# You could rewrite this code as:
  efa_model <- '
              efa("block1")*F1 + efa("block1")*F2 =~ V1 + V2 + V3 + V4 + V5 + V6
              '
```

# Estimate the Model

```{r}
efa_f2 <- 
  sem(model = efa_model,
      data = efa_data,
      rotation = "oblimin",
      estimator = "MLR"
      )


```

We got a warning: 

Warning: lavaan->lav_object_post_check():  
   some estimated ov variances are negative
   
We'll ignore this for now, but this is not ignorable (variables cannot be negative).
   
# Request the Output

```{r}
summary(efa_f2, 
        fit.measures = TRUE, 
        standardized = TRUE
        )
```

```{r}

# lavInspect() is a helpful function to get additional informaiton on the model

# Here, you can see the parameter numbers (as in Mplus' TECH1)
lavInspect(efa_f2)

# By asking for "sampstat" you can see the observed var/covar matrix (as in Mplus)
lavInspect(efa_f2,"sampstat")

# There are MANY other options. To see residual var/covar matrix:
lavInspect(efa_f2,"resid")

# To see R2 values (note there's an NA for V1 because it's fixed to 1 by default)
lavInspect(efa_f2,"rsquare") # the proportion of variance explained (akin to R-squared) for each observed variable in your model.

#... And lots more

```

## You can also run an EFA using the cfa() function of lavaan, by specifying auto.efa = TRUE

```{r}
efa_f2 <- 
  cfa(model = efa_model,
      data = efa_data,
      rotation = "oblimin",
      estimator = "MLR",
      auto.efa = TRUE
      )
```

# Table

```{r}
  summary(efa_f2, 
          fit.measures = TRUE, 
          standardized = TRUE
          )

  
  library(lavaanExtra)
  
  fit_table <- nice_fit(efa_f2, nice_table = TRUE)
  
```

# Plot

```{r}
  
  library(semPlot)
  semPaths(efa_f2, what = "std", layout = "tree")
  
  
  
  library(lavaanPlot)
  lavaanPlot(
  model = efa_f2,
  coefs = TRUE,
  covs = FALSE,
  stand = TRUE,
  digits = 2
)

```
