---
title: "W3 practice"
output:
  pdf_document: default
  html_document: default
date: '2023-01-26'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


##########################################
############# week 3 exercise ############
##########################################
library(haven); library(psych); library(dplyr); 
library(magrittr); library(ggplot2); library(gridExtra)
library(epitools); library(lsr); library(descr); library(epiR); library(epiDisplay)

```


# 0. data step 
```{r, echo = FALSE, message = FALSE, warning = FALSE}

library(haven); library(psych); library(dplyr); 
library(magrittr); library(ggplot2); library(gridExtra)
library(epitools); library(lsr); library(descr); library(epiR); library(epiDisplay)

  dat <- read_dta("w3/Choices.dta")
  names(dat) # name of the variables
  dat %>% dim # 2536, 16
  summary(dat) # get min, max, NA's
  describe(dat)
```  

# 1. Create a new variable (pref2cat) 

```{r}
  dat$pref %>% hist
  dat$gender
  dat$pref %>% median()
  
  
  dat$gender %>% table()

  dat = 
    dat %>% 
    mutate(pref2cat = ifelse(pref > 1, 1, 0),
           gender.ch = ifelse(gender == 1, "F", "M"),
           race.ch = ifelse(race == 1, "W", "B"))
```

# 2. Define the percentage of Black men and Black women

```{r}

  tab = table(dat$race, dat$gender)
  tab[1,2]/sum(tab)*100 # percentage of black man
  tab[1,1]/sum(tab)*100 # percentage of blakc women
```


# 3. Calculate the difference in proportion between men and women for prefcat=1 vs prefcat=0

```{r}

  ## with a 95% confidence interval (Risk/prevalence difference)
  cont = xtabs( ~ gender + pref2cat, dat = dat) # create a contingency table
  riskratio(cont) # difference in proportion: 434/1556*100 - 292/980*100
  riskratio(dat$gender, dat$pref2cat) # this works too

  prop.test(x = c(434, 292), n = c(1556, 980))

```

# 4. Conduct a chi square test for with gender and prefcat
```{r}

  chisq.test(cont)
```

# 5. Maental-Haenzel stratified analysis
```{r}


  mhor(dat$gender, dat$pref2cat, dat$race)
```

# 6. Conduct an independent samples t-test comparing the mean preference score between men and women
```{r}

  t.test(pref ~ gender, data = dat) # outcome ~ predictor
```

# 7. Conduct a paired t-test comparing pref and fpref in the dataset

```{r}

  # t.test(before, after, paired = TRUE)
  t.test(dat$fpref, dat$pref, paired = TRUE)
```

# 8. anova

```{r}
dat$educ %>% class
dat$educ = dat$educ %>% as.factor()

# dat =
#   dat %>%
#   mutate(educ = as.factor(educ))

  fit = lm(pref ~ educ, data = dat)
  anova(fit)
  pairwise.t.test(dat$pref, dat$educ, p.adjust.method="bonferroni")


```

