---
title: "Complex Measurement Models - MTMM Test for Construct Validity - CU"
subtitle: "Applied SEM Lab (PSY 768)" 
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
options(scipen=999)
```

```{r Libraries, warnings = F, message = F, echo = FALSE}
library(lavaan)
library(readr)
library(semTools)
```
**This demo is based on data and models from Byrne, B.M. (2011). Structural Equation Modeling with Mplus: Basic Concepts, Applications, and Programming (1st ed.). Routledge. https://doi.org/10.4324/9780203807644**

## Correlated Uniquenesses (CU)

```{r}
mtmm_data <- read.table("ind7mt.dat")
```

```{r}
colnames(mtmm_data) <- c(
  'scself',
  'sctch',
  'scpar',
  'scpeer',
  'acself',
  'actch',
  'acpar',
  'acpeer',
  'ecself',
  'ectch',
  'ecpar',
  'ecpeer',
  'mcself',
  'mctch',
  'mcpar',
  'mcpeer'
)
```

## Model Specification

This model only embodies the 4 correlated traits. We do not specify method effects per se. That is, the CU does not assume that method effects associated with a certain method can be explained by a single latent method factor. 

However, their effects are implied by specifying correlated uniquenesses (residuals) associated with each set of observed variables with same method.

```{r}
mtmm_cu <- '
# Latent factors
# F1-F4 are latent methods (parent, teacher, self, peer report)
F1 =~ NA*scself + sctch + scpar + scpeer
F2 =~ NA*acself + actch + acpar + acpeer
F3 =~ NA*ecself + ectch + ecpar + ecpeer
F4 =~ NA*mcself + mctch + mcpar + mcpeer

# F5-F8 are latent traits (social, academic, english, math competence)
# Note that we are removing these from the model and specifying
# the correlated uniquenesses instead
# F5 =~ NA*scself + acself + ecself + mcself
# F6 =~ NA*sctch + actch + ectch + mctch 
# F7 =~ NA*scpar + acpar + ecpar + mcpar
# F8 =~ NA*scpeer + acpeer + ecpeer + mcpeer

# Latent variable variances fixed to 1
F1 ~~ 1*F1
F2 ~~ 1*F2
F3 ~~ 1*F3
F4 ~~ 1*F4

# Covariances among the residuals of the methods
scself ~~ acself
scself ~~ ecself 
scself ~~ mcself
acself ~~ ecself
acself ~~ mcself
ecself ~~ mcself
sctch ~~ actch 
sctch ~~ ectch 
sctch ~~ mctch
actch ~~ ectch
actch ~~ mctch
ectch ~~ mctch
scpar ~~ acpar 
scpar ~~ ecpar 
scpar ~~ mcpar
acpar ~~ ecpar 
acpar ~~ mcpar
ecpar ~~ mcpar
scpeer ~~ acpeer 
scpeer ~~ ecpeer 
scpeer ~~ mcpeer
acpeer ~~ ecpeer 
acpeer ~~ mcpeer
ecpeer ~~ mcpeer
'
```

## Fitting the model

```{r}

mtmm_cu_fit <- cfa(model = mtmm_cu,
               data = mtmm_data,
               missing = "ml",
               estimator = "mlr",
               control = list(iter.max = 1000)
               )

```

Note that fit is very good and we didn't need to fix the variance of SCSELF to 0. 

## Model Output 

### Model summary

```{r}
summary(mtmm_cu_fit, 
        standardized = TRUE, 
        rsquare = TRUE,
        fit.measures = TRUE)
```

