---
title: "Latent Growth Model in lavaan"
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r Packages}
library(psych)
library(dplyr)
library(lavaan)
library(ggplot2)
```

## Load the Dataset

```{r Loading files}
hkcancer <- read.table("hkcancer_red2.dat")
hkcancer <- data.frame(lapply(hkcancer, function(x) as.numeric(as.character(x))))
```

## Add variable names

```{r}
colnames(hkcancer) <- c('AGE_GP',
                       'SURG_TX',
                       'mood2',
                       'mood3',
                       'mood4',
                       'socadj2', 
                       'socadj3', 
                       'socadj4',
                       'age',
                       'age2', 
                       'surg_tx2'
                        )
```

# Latent Curve Models

## Missing data patterns

We'll first check our missing data patterns using the ```mdpattern()``` function of the *mice* package.

The number of rows in each of the output windows indicates the number of missing data pattenrs. The second window, with the blue and red graphic, makes it easier to visualize the patterns. The first row, with all blue columns refer to the patter in which we have no missing data (307 cases).

We can also see that 12 out of 405 were missing the mood and social adjustment outcome at the third time point (i.e., missing MOOD4 & SOCADJ4). This information is helpful for reporting attrition and missingness in your data in a write-up.

```{r}
library(mice)
md.pattern(hkcancer)
```

```{r}
describe(hkcancer)
```

## Model 0 - intercept only model

We'll start with an intercept only model, which assumes that people may vary in their starting point but continue at the same level over time - rate of change is 0 for everyone
 
```{r}

Model0 <- "
           # Intercept and slope of mood and social adjustment
           I1 =~ 1*mood2 + 1*mood3 + 1*mood4
           I2 =~ 1*socadj2 + 1*socadj3 + 1*socadj4
           
           # intercept of observed variables fixed to 0 
           mood2 + mood3 + mood4 ~ 0*1
           socadj2 + socadj3 + socadj4 ~ 0*1
           
           # variances
           I1 ~~ I1
           I2 ~~ I2
          "
```

## Fit the model

```{r}
Model0.fit <-
  growth(
    model = Model0,
    data = hkcancer,
    missing = "FIML",
    estimator = "MLR",
    meanstructure = TRUE
  )
```

## Request the output

Fit is inadequate in terms of RMSEA, SRMR and CFI. 

```{r}
summary(Model0.fit, 
        fit.measures = T, 
        standardized = T)
```

Although fit is pretty good, notice that trajectories over time don't seem to be flat, on average 

# The data, on average

```{r}
sum_data <-'adjustment =~ socadj2 + socadj3 + socadj4'
sum_data.fit <- sem(model = sum_data, data = hkcancer, missing = "ML")
mean_adj <- (lavInspect(sum_data.fit, what="mu.hat"))
time <- seq(from = 1, to = 3, by = 1)
plot(x = time, y = mean_adj, xlab = "Time", ylab = "Social adjustment")
lines(x = time, y = mean_adj)
```

```{r}
mood_data <-'adjustment =~ mood2 + mood3 + mood4'
mood_data.fit <- sem(model = mood_data, data = hkcancer, missing = "ML")
mean_mood <- (lavInspect(mood_data.fit, what="mu.hat"))
time <- seq(from = 1, to = 3, by = 1)
plot(x = time, y = mean_mood, xlab = "Time", ylab = "Mood")
lines(x = time, y = mean_mood)
```
And if we decompose these trajectories into individual ones, we can see that this is really not the case. 

# The data, by participant

```{r}
library(lcsm)

hkcancer <- dplyr::mutate(hkcancer, ID = row_number())

plot_trajectories(
  hkcancer,
  id_var = "ID",
  var_list = c("mood2", "mood3", "mood4"),
  line_colour = "blue",
  xlab = "Time",
  ylab = "Mood",
  scale_x_num = TRUE,
  scale_x_num_start = 1,
  connect_missing = TRUE)

```
```{r}
plot_trajectories(
  hkcancer,
  id_var = "ID",
  var_list = c("socadj2", "socadj3", "socadj4"),
  line_colour = "blue",
  xlab = "Time",
  ylab = "Soc Adj",
  scale_x_num = TRUE,
  scale_x_num_start = 1,
  connect_missing = TRUE)
```

## Model 1

We can specify LGM in lavaan with both intercept and slope in the same way we do in Mplus: by creating an intercept and a slope latent factor. 

This measurement model is basically a factor analysis with 2 special features: (1) all loadings are fixed; (2) pattern of factor loadings & mean structure allows us to interpret as slope & intercept factors.

We'll end up with matrices/vectors which include: (1) constants & known time values invariant across individuals; (2) individual growth parameters (w/in person true change); (3) residuals of outcomes (w/in person noise).

```{r}

Model1 <- "
           # Intercept and slope of mood and social adjustment
           I1 =~ 1*mood2 + 1*mood3 + 1*mood4
           S1 =~ 0*mood2 + 1*mood3 + 2.33*mood4
           I2 =~ 1*socadj2 + 1*socadj3 + 1*socadj4
           S2 =~ 0*socadj2 + 1*socadj3 + 2.33*socadj4
           
           # factor covariances
           I1 ~~ S1
           I2 ~~ S2
           
           # intercept of observed variables fixed to 0 
           mood2 + mood3 + mood4 ~ 0*1
           socadj2 + socadj3 + socadj4 ~ 0*1
           
           # variances
           I1 ~~ I1
           S1 ~~ S1
           I2 ~~ I2
           S2 ~~ S2
          "
```

Why the weird loadings for time? Recall that these data were collected 1, 4, and 8 months post-sugery. A change in loading from 0-1 refers to 3 months (4 months minus 1 month). A change in loading from 1-2.33 refers to 4 months (8 months minus 4 months). If using LGM, you will need to ensure that the spacing between assessments is proportionate to the true time between assessments.

## Fit the model

```{r}
Model1.fit <-
  growth(
    model = Model1,
    data = hkcancer,
    missing = "FIML",
    estimator = "MLR",
    meanstructure = TRUE
  )
```

## Request the output

Fit is inadequate in terms of RMSEA, SRMR and CFI. 

```{r}
summary(Model1.fit, 
        fit.measures = T, 
        standardized = T)
```

## Interpret the output

Let's start with fit. Because intercept-only and intercept+slope models are nested within each other, we can use likelihood ratio test to assess change in model fit. However, we estimated these models using MLR, so we need to use the scaling correction factor in the LRT. We'll cut a corner here and use the anova() function of lavaan, which does that for us automatically. 

As you can see, the fit significantly improves when we add a slope factor to the model. We'll stop here, but we could have also tested whether residuals should be homo- or heteroskedastic. If we had more time points, we could also test if adding a quadratic slope improves fit.

```{r}
lavaan::anova(Model0.fit,Model1.fit)
```

Now, to the coefficients: 

The first part of the output refers to the factor loadings, which we fixed to create the intercept and the slope factors. The second part of the output refers to the covariances. 

We'll first look at the intercepts.

```
Intercepts:
         Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all

    I1    21.450    0.318   67.440    0.000    4.228    4.228
    S1    -0.572    0.133   -4.297    0.000   -0.381   -0.381
    I2    100.865   0.506  199.250    0.000   11.331   11.331
    S2    -0.255    0.208   -1.229    0.219   -0.079   -0.079
```

The intercepts of the growth parameters (intercept and slope) refer to the average scores of self-reported mood and social adjustment at the first time point in the model (i.e., where the slope was fixed to 0, which is at month 1 - this could be changed to any month) and the average rate of change across the study period (8 months).

The intercept of I1 means that the average self-reported mood at 1 month was 21.450 (here, lower value = better mood).

The intercept of S1 indicates the mood score decreased by .57, on average, across the eight months. In other words, as time since surgery increased, there was a gradual increase in positive mood.

The intercept of I2 indicates that the average level of social adjustment at 1 month was 100.87.

The slope of S2 is not significant, which means that there was negligible (n.s.) change in social adjustment over the 8-month time period.

We should also interpret the variances of the growth factors.

```
Variances:
             Estimate  Std.Err  z-value  P(>|z|)   Std.lv  Std.all
    I1         25.741    3.694    6.968    0.000    1.000    1.000
    S1          2.254    1.286    1.752    0.080    1.000    1.000
    I2         79.236   12.666    6.256    0.000    1.000    1.000
    S2         10.339    3.531    2.928    0.003    1.000    1.000
```

The variance of I1 means that there are significant inter-individual differences in initial status on mood.

The variance of S1 means that there are non-significant inter-individual slope (change) in mood.

The variance of I2 means that there are significant inter-individual differences in initial status on social adjustment.

!The variance of S2 means that there are significant inter-individual differences in slope (change) in social adjustment.

Because we have significant inter-individual differences, we want to try to explain those in subsequent analyses with predictors. That is, what is contributing to these differences? To do that, we'll add predictors to our model.

The covariance between the slope and intercept of mood is not significant, which suggests that participants' average mood is not related to the extent to which their mood changes over time.

```
                   Estimate  Std.Err  z-value  P(>|z|)
  I1 ~~                                               
    S1               -2.013    1.666   -1.209    0.227
```

The covariance between the slope and intercept of social adjustment is significant. This suggests that participants' average level of social adjustment is  related to the extent to which their social adjustment changes over time. Because this relationship is negative, it indicates that participants generally higher in social adjustment tend to have more stable (less changeable) social adjustment over time.

```
                   Estimate  Std.Err  z-value  P(>|z|)
  I2 ~~                                               
    S2              -13.386    4.634   -2.889    0.004
```

There's also a significant covariance between the intercept of mood and social adjustment, such that people higher in mood also tend to be higher in social adjustment.

```
                   Estimate  Std.Err  z-value  P(>|z|)
  I1 ~~                                               
    I2              -23.395    4.007   -5.839    0.000
```

Finally, the following covariances are non-significant (in order): 
1) the covariance between the intercept of mood and the variance of social adjustment; 
2) the covariance between the slope of mood and the intercept of social adjustment
3) the slope of mood and the slope of social adjustment (p = .046; significant, but barely!).

# Plotting the predicted trajectories

```{r}

predicted.df <- lavPredict(Model1.fit, type = "lv", method = "EBM",
label = T)

predicted.df2 <- data.frame(hkcancer[, c(12, 3:8)], predicted.df)

# Preview the first 10 lines:
head(predicted.df2, n = 10)

predicted.df3 <- reshape(data = predicted.df2, varying = c("mood2", "mood3", "mood4", "socadj2","socadj3", "socadj4"),
timevar = "time", idvar = "ID", direction = "long", sep = "")

predicted.df3$time = predicted.df3$time - 1
predicted.df3 <- predicted.df3[order(predicted.df3$ID, predicted.df3$time), ]

predicted.df3["pred"] <- predicted.df3$I1 + predicted.df3$S1 * predicted.df3$time

plot1 <- ggplot(predicted.df3[predicted.df3$ID <= 100,], aes(x = time, y = pred, group = ID))
plot1 <- plot1 + theme_classic(base_family = "sans") + 
  theme(text = element_text(size = 12)) + 
  labs(x = "Time", y = "Mood") +
#scale_x_continuous(limits = c(0, 7), breaks = c(0:7)) +
#scale_y_continuous(limits = c(0, 10), breaks = c(0:10)) +
  geom_line(size = 0.2, colour = "#000000")
  plot1
  
```

```{r}
predicted.df3["pred"] <- predicted.df3$I2 + predicted.df3$S2 * predicted.df3$time

plot1 <- ggplot(predicted.df3[predicted.df3$ID <= 100,], aes(x = time, y = pred, group = ID))
plot1 <- plot1 + theme_classic(base_family = "sans") + 
  theme(text = element_text(size = 12)) + 
  labs(x = "Time", y = "Social Adjustment") +
#scale_x_continuous(limits = c(0, 7), breaks = c(0:7)) +
#scale_y_continuous(limits = c(0, 10), breaks = c(0:10)) +
  geom_line(size = 0.2, colour = "#000000")
  plot1
```

## Adding covariates to the model

```{r}
covModel <- "
           # slopes and intercepts
           I1 =~ 1*mood2 + 1*mood3 + 1*mood4
           S1 =~ 0*mood2 + 1*mood3 + 2.33*mood4
           I2 =~ 1*socadj2 + 1*socadj3 + 1*socadj4
           S2 =~ 0*socadj2 + 1*socadj3 + 2.33*socadj4

           # intercepts fixed to 0
           mood2 + mood3 + mood4 ~ 0*1
           socadj2 + socadj3 + socadj4 ~ 0*1
           
           # variances of the growth parameters
           I1 ~~ I1
           S1 ~~ S1
           I2 ~~ I2
           S2 ~~ S2
           
           # regression paths
           I1 + S1 ~ age2 + surg_tx2
           I2 + S2 ~ age2 + surg_tx2
          "

```

## Fit the model

```{r}
covModel.fit <-
  growth(
    model = covModel,
    data = hkcancer,
    missing = "FIML",
    estimator = "MLR",
    meanstructure = TRUE
    )
```

## Request the output

```{r}
summary(
  covModel.fit,
  fit.measures = T,
  standardized = T,
  rsquare = T
)
```


Fit is good when we include the predictors/covariates in the model.

All parts of the output are the same, except that now we also have regression parameters. 

```
                   Estimate  Std.Err  z-value  P(>|z|)  
  I1 ~                                                         
    age2             -1.604    0.626   -2.564    0.010   
    surg_tx2         -0.123    0.757   -0.163    0.870  
```

This output indicates that age is a significant predictor of initial mood, such that mood scores are lower for older women at time 1 (lower mood scores = more positive mood). However, mood scores didn't vary  based on type of surgery.

```
                   Estimate  Std.Err  z-value  P(>|z|)  
  S1 ~                                                         
    age2             -0.214    0.261   -0.822    0.411   
    surg_tx2         -0.326    0.336   -0.971    0.332  
```

Here we see that the differences in slopes (i.e., rate of change) of mood were not significant across age groups or surgery type.

```
  I2 ~                                                         
    age2             -0.963    0.989   -0.974    0.330   
    surg_tx2         -3.242    1.052   -3.081    0.002   
```
The first line above indicates that age is not a significant predictor of initial status on soc adjustment. The second line indicates that surgery type *is* a significant predictor of initial status on social adjustment, such that those undergoing a mastectomy had lower social adjustment than those not undergoing a mastectomy.

```
                   Estimate  Std.Err  z-value  P(>|z|)   
  S2 ~                                                         
    age2              0.368    0.406    0.907    0.365    
    surg_tx2          1.239    0.487    2.545    0.011
```

The first line of the output above indicates that age is not a significant predictor of slopes (i.e., rate of change) on social adjustment.
The second line indicates that surgery type *is* a significant predictor. Those undergoing mastectomy had faster rate of change in social adjustment relative to other surgery.

## Different location for t = 0?

```{r}
covModel2 <- "
           # slopes and intercepts
           I1 =~ 1*mood2 + 1*mood3 + 1*mood4
           S1 =~ -1.75*mood2 + -1*mood3 + 0*mood4
           I2 =~ 1*socadj2 + 1*socadj3 + 1*socadj4
           S2 =~ -1.75*socadj2 + -1*socadj3 + 0*socadj4

           # intercepts fixed to 0
           mood2 + mood3 + mood4 ~ 0*1
           socadj2 + socadj3 + socadj4 ~ 0*1
           
           # variances of the growth parameters
           I1 ~~ I1
           S1 ~~ S1
           I2 ~~ I2
           S2 ~~ S2
           
           # regression paths
           I1 + S1 ~ age2 + surg_tx2
           I2 + S2 ~ age2 + surg_tx2
          "

```

## Fit the model

```{r}
covModel.fit2 <-
  growth(
    model = covModel2,
    data = hkcancer,
    missing = "FIML",
    estimator = "MLR",
    meanstructure = TRUE
    )
```

## Request the output

```{r}
summary(
  covModel.fit2,
  fit.measures = T,
  standardized = T,
  rsquare = T
)
```

## Different location for t = 0 and t = 1?

```{r}
covModel3 <- "
           # slopes and intercepts
           I1 =~ 1*mood2 + 1*mood3 + 1*mood4
           S1 =~ 0*mood2 + 0.42857143*mood3 + 1*mood4
           I2 =~ 1*socadj2 + 1*socadj3 + 1*socadj4
           S2 =~ 0*socadj2 + 0.42857143*socadj3 + 1*socadj4

           # intercepts fixed to 0
           mood2 + mood3 + mood4 ~ 0*1
           socadj2 + socadj3 + socadj4 ~ 0*1
           
           # variances of the growth parameters
           I1 ~~ I1
           S1 ~~ S1
           I2 ~~ I2
           S2 ~~ S2
           
           # regression paths
           I1 + S1 ~ age2 + surg_tx2
           I2 + S2 ~ age2 + surg_tx2
          "

```

## Fit the model

```{r}
covModel.fit3 <-
  growth(
    model = covModel3,
    data = hkcancer,
    missing = "FIML",
    estimator = "MLR",
    meanstructure = TRUE
    )
```

## Request the output

```{r}
summary(
  covModel.fit3,
  fit.measures = T,
  standardized = T,
  rsquare = T
)
```
## Change in specification of time?

```{r}
covModel4 <- "
           # slopes and intercepts
           I1 =~ 1*mood2 + 1*mood3 + 1*mood4
           S1 =~ 0*mood2 + 3*mood3 + 7*mood4
           I2 =~ 1*socadj2 + 1*socadj3 + 1*socadj4
           S2 =~ 0*socadj2 + 3*socadj3 + 7*socadj4

           # intercepts fixed to 0
           mood2 + mood3 + mood4 ~ 0*1
           socadj2 + socadj3 + socadj4 ~ 0*1
           
           # variances of the growth parameters
           I1 ~~ I1
           S1 ~~ S1
           I2 ~~ I2
           S2 ~~ S2
           
           # regression paths
           I1 + S1 ~ age2 + surg_tx2
           I2 + S2 ~ age2 + surg_tx2
          "

```

## Fit the model

```{r}
covModel.fit4 <-
  growth(
    model = covModel4,
    data = hkcancer,
    missing = "FIML",
    estimator = "MLR",
    meanstructure = TRUE
    )
```

## Request the output

```{r}
summary(
  covModel.fit4,
  fit.measures = T,
  standardized = T,
  rsquare = T
)
```